---
title: Transcriptome Dimension Reduction
output:
  html_document:
        toc: yes
---


```{r, echo=FALSE}
## input/output setup
io <- list()
io$chunk$cache <- !TRUE
io$chunk$echo <-FALSE

knitr::opts_chunk$set(cache=io$chunk$cache, echo=io$chunk$echo,  cache.comment=FALSE, fig.align = 'center',
                      message=FALSE, warning=FALSE, results='hide')
```

```{r, cache=FALSE}
## source directory
source_dir <- '../../utils'
R_files <- list.files(source_dir, pattern = '.R$', full.names = TRUE) ## all files ending in .R
invisible(lapply(R_files, base::source))

io$input$rna_sce_normalised <- '../../../output/parsed/rna_sce_normalised.Rds'
## check that files exist
check_files_exist(io$input)
io$opts$top_hvgs = 500
io$opts$pcs = 5
## load
rna_sce <- readRDS(io$input$rna_sce_normalised)
```
```{r}
## veriance decomosition
rna_sce %<>% sce_var_decomp()
```

## Variance vs Mean Expression

```{r}
plot_trendvar(rna_sce)
```

## All Genes

### PCA

```{r}
## pca
pca_res_all = pca(t(logcounts(rna_sce)), ncomp = io$opts$pcs)
```
```{r, out.width="100%"}
## plot
pca_grid(pca_obj = pca_res_all, top = 'PCA plots - all genes')
```

### Hierarchical Clustering

```{r, cache=TRUE}
plotHeatmap(rna_sce, features = rownames(rna_sce), main = paste0( 'Clustered heatmap of scRNAseq data \n cells at the bottom and genes on the side'))
```

## Top HVGs - Total Variance

Dimension reduction using top `r io$opts$top_hvgs` hvgs based on total variance:

```{r}
rna_sce_hvg_total = sce_subset(rna_sce, row_data = 'total',
                                          n_top = io$opts$top_hvgs)
```

### PCA

```{r}
## pca
pca_res_hvgs_total = pca(t(logcounts(rna_sce_hvg_total)), ncomp = io$opts$pcs)
```
```{r, out.width="100%"}
## plot
pca_grid(pca_obj = pca_res_hvgs_total, top = 'PCA plots - top HVGs - total variance')
```


### Hierarchical Clustering

```{r, cache=TRUE}
rna_sce_hvg_total %>%  plotHeatmap(., features = rownames(.), main = paste0( 'Clustered heatmap of scRNAseq data using HVGs based on total variance'))
```

## Top HVGs - Decomposed Biological Variance

Dimension reduction using top `r io$opts$top_hvgs` hvgs based on decomposed biological variance:

```{r}
rna_sce_hvg_bio = sce_subset(rna_sce, row_data = 'bio',
                                          n_top = io$opts$top_hvgs)
```

### PCA

```{r}
## pca
pca_res_hvgs_bio = pca(t(logcounts(rna_sce_hvg_bio)), ncomp = io$opts$pcs)
```
```{r, out.width="100%"}
## plot
pca_grid(pca_obj = pca_res_hvgs_bio, top = 'PCA plots - top HVGs - bio. variance')
```

### Hierarchical Clustering

```{r, cache=TRUE}
rna_sce_hvg_bio %>%  plotHeatmap(., features = rownames(.), main = paste0( 'Clustered heatmap of scRNAseq data using HVGs based on bio. variance'))
```

