---
title: Transcriptome QC and Normalisation - all cells
output:
  html_document:
        toc: yes
---

```{r, echo=FALSE}
## input/output setup
io <- list()
io$chunk$cache <- !TRUE
io$chunk$echo <-FALSE

knitr::opts_chunk$set(cache=io$chunk$cache, echo=io$chunk$echo,  cache.comment=FALSE, fig.align = 'center',
                      message=FALSE, warning=FALSE, results='hide')
```

```{r, cache=FALSE}
## source directory
source_dir <- '../../utils'
R_files <- list.files(source_dir, pattern = '.R$', full.names = TRUE) ## all files ending in .R
invisible(lapply(R_files, base::source))

source("../../all_libs.R")
io$opts$min_cells_expr <- 0.25 ## homogeneous population
io$opts$min_log10exp <- 0.7 ## minimum log expression based on lib size histpgram
io$input$rna_raw <- '../../../data/scRNAseq/raw/Kasumi1_scRNAseq_RawCounts.txt'
io$output$rna_sce_normalised_all_cells <- '../../../output/parsed/rna_sce_normalised_all_cells.Rds'
## check that files exist
check_files_exist(io$input)
```

<!-- ## Load Raw Data -->

```{r tablecounts}
## read raw scRNAseq counts
counts_raw <- read.csv(io$input$rna_raw, row.names = 1)
## change the cell names based on well names
colnames(counts_raw) <- name_cells_by_wells(colnames(counts_raw), underscores = c(1,2))
## create sce object from raw counts
rna_sce <- SingleCellExperiment(assays = list(counts = as.matrix(counts_raw)))
```

```{r qc}
## add QC metrics
rna_sce <- scater::calculateQCMetrics(rna_sce)
## filter undetected genes and add percent of genes detected in colData
rna_sce <- sce_detected_genes(rna_sce)
```


## Gene QC

To filter genes with high level of dropouts, genes to have a log mean expression of at least `r io$opts$min_log10exp`:

```{r}
## gene QC
gene_metrics <- as.data.frame(rowData(rna_sce)) ## a data.frame of gene metrics

## filter the first peak of non-zero values
hist <- hist(gene_metrics$log10_mean_counts, prob=TRUE, breaks = 100, col = 'grey80',
            main= 'Histogram of log10 mean counts', xlab = 'log10[mean counts] for each gene')
abline(v=io$opts$min_log10exp, col='red', lwd=2)
```
```{r}
rna_sce %<>% .[rowData(.)$log10_mean_counts>=io$opts$min_log10exp,]
```

```{r, echo=TRUE, results='show'}
dim(rna_sce)
```

```{r}
## recalculate QC metrics
rna_sce <- scater::calculateQCMetrics(rna_sce)
rna_sce <- sce_detected_genes(rna_sce)
```

## Cell QC

Cells to express at least `r io$opts$min_cells_expr*100` percent of genes:

```{r}
cell_qc_out <- cell_qc(sce = rna_sce, pct_genes_in_cells = 100*io$opts$min_cells_expr)
cell_qc_out$plot
```

`r ifelse(dim(rna_sce)[2]==dim(cell_qc_out$sce)[2], paste0('No cell filtering needed. Minimum percent of genes expressed: ', round(min(rna_sce$pct_total_features_by_counts)),'% .'), paste0('Number of cells that passed filtering: ', dim(cell_qc_out$sce)[2],'.'))`

```{r}
rna_sce <- cell_qc_out$sce
```

```{r, echo=TRUE, results='show'}
dim(rna_sce)
```
```{r}
## normalise using scran
rna_sce <- computeSumFactors(rna_sce) ## add size factors
rna_sce <- normalize(rna_sce) ## normalise
```

## Size Factors

```{r}
plot(sizeFactors(rna_sce), rna_sce$total_counts/1e6, log='xy', pch=16, col='purple',
    ylab='Library size (x 1e6)', xlab='Size factor', main = 'Scran Size Factors')
```
```{r}
saveRDS(rna_sce, file=io$output$rna_sce_normalised_all_cells)
```

